ForOpenCL: Transformations Exploiting Array Syntax in Fortran for Accelerator Programming

### 2011-12-11 papers/IJCSE/intro.tex
edits to intro

A directive-based approach, such as OpenMP or the Accelerator programming model
from the Portland Group~\citep{pgi10accelerator}, is one solution to this
problem.  However, in this paper we take a somewhat different approach.  A
common theme amongst the new processors is the emphasis on data-parallel
programming.  This model is well-suited to architectures that are
based on either vector processing or massively parallel collections of simple
cores.  The recent CUDA and OpenCL programming languages are intended to
support this programming model.

@@

We demonstrate that a subset of Fortran map surprisingly well onto GPUs when
transformed to OpenCL kernels.  This data-parallel subset includes: array syntax
using assignment statements and binary operators, array constructs like {\tt
  WHERE}, and the use of pure and elemental functions.  In addition, we provide
new functions that explicitly take advantage of the stencil geometry of the
problem domain we consider.  Note that this subset of the Fortran language is
implicitly parallel.  This programming model \emph{does not require explicit
  declaration of parallelism within the program.}  In addition, programs are
expressed using entirely standard Fortran so it can be compiled for and executed
on a single core without concurrency.

>>>>

A directive-based approach, such as OpenMP or the Accelerator
programming model from the Portland Group~\citep{pgi10accelerator} and
now generalized in the OpenACC standard~\citep{openacc11std}, is one
solution to this problem.  However, in this paper we take a somewhat
different approach not requiring explicit declaration of parallelism,
instead relying on implicitly parallel subsets of the Fortran
language.  A common theme amongst the new processors is the emphasis
on data-parallel programming.  This model is well-suited to
architectures that are based on either vector processing or massively
parallel collections of simple cores.  The recent CUDA and OpenCL
programming languages are intended to support this programming model.

@@

We demonstrate that a subset of Fortran maps surprisingly well onto
GPUs when transformed to OpenCL kernels.  This data-parallel subset
includes: array syntax using assignment statements and binary
operators, array constructs like {\tt WHERE}, and the use of pure and
elemental functions.  In addition, we provide new functions that
explicitly take advantage of the stencil geometry of the problem
domain we consider.  Note that this subset of the Fortran language is
implicitly parallel.  This programming model \emph{does not require
  explicit declaration of parallelism within the program.}  In
addition, programs are expressed using entirely standard Fortran so it
can be compiled for and executed on a single core without concurrency.
